---
title: "R vs  Python: Different similarities and similar differences"
author: "Tinniam V Ganesh"
date: "May 18, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R versus Python 
A debate about which  language is better suited for Datascience, R or Python, can set off diehard fans of these languages into a tizzy. This post tries to look at some of the different similarities and similar differences between these languages. To a large extent the ease or difficulty in learning R or Python is subjective. I have heard that R has a steeper learning curve than Python and also vice versa. This probably depends on the degree of familiarity with the languuge
To a large extent both R an Python do the same thing in just slightly different ways and syntaxes. The ease or the difficulty in the R/Python construct's largely is in the 'eyes of the beholder' nay, programmer' we could say. I include my own experience with the languages below.

###1. R data types
R has 2 data types

a) Character 
b) Numeric

Python has several data types

a) Int
b) float
c) Long
d) Complex
and so on

- R has the type vector which in Python is closely related to a list
- To know the type of the variable we use the following for R and Python
- R & Python use length & len respectively to get the length of a vector and list respectively

###2. R Vector vs Python List
A common data type in R is the vector. Python has a similar data type, the list

```{r cache=TRUE}
# R vectors
a<-c(4,5,1,3,4,5)
print(a[3])
print(a[3:4]) # R does not always need the explicit print. 
#R type of variable
print(class(a))
# Length of a
print(length(a))
```

```{python cache=TRUE}
# Python lists
a=[4,5,1,3,4,5] # 
print(a[2]) # Some python IDEs require the explicit print
print(a[2:5])
print(type(a))
# Length of a
print(len(a))
```

###2a. Other data types - Python
Python also has certain other data types like the tuple, dictionary etc as shown below. R does not have as many of the data types, nevertheless we can do everything that Python does in R

```{python cache=TRUE}
# Python tuple
b = (4,5,7,8)
print(b)


#Python dictionary
c={'name':'Ganesh','age':54,'Work':'Professional'}
print(c)
#Print type of variable c

```


###2.Type of Variable 
To know the type of the variable in R we use 'class', In Python the corresponding command
is 'type'

```{r cache=TRUE}
#R - Type of variable
a<-c(4,5,1,3,4,5)
print(class(a))

```

```{python cache=TRUE}
#Python - Print type of tuple a
a=[4,5,1,3,4,5]
print(type(a))
b=(4,3,"the",2)
print(type(b))
```

###3. Length 
To know length in R, use length()
```{r cache=TRUE}
#R - Length of vector
# Length of a
a<-c(4,5,1,3,4,5)
print(length(a))
```

To know the length of a list,tuple or dict we can use len()
```{python cache=TRUE}
# Python - Length of list , tuple etc
# Length of a
a=[4,5,1,3,4,5]
print(len(a))
# Length of b
b = (4,5,7,8)
print(len(b))

```


###4. Accessing help

To access help in R we use the '?' or the 'help' function 
```{r cache=TRUE}
#R - Help - To be done in R console or RStudio
#?sapply
#help(sapply)
```

Help in python on any topic involves
```{python cache=TRUE}
#Python help - This can be done on a (I)Python console
#help(len)
#?len
```

###5. Subsetting
The key difference between R and Python with regards to subsetting is that in R the index starts at 1. In Python it starts at 0, much like C,C++ or Java
To subset a vector in R we use
```{r cache=TRUE}
#R - Subset
a<-c(4,5,1,3,4,8,12,18,1)
print(a[3])

# To print a range or a slice. Print from the 3rd to the 5th element
print(a[3:6])
```

Python also uses indices. The difference in Python is that the index starts from 0/
```{python cache=TRUE}
#Python - Subset
a=[4,5,1,3,4,8,12,18,1]
# Print the 4th element (starts from 0)
print(a[3])

# Print a slice from 4 to 6th element
print(a[3:6])
```

###6. Operations on vectors in R and operation on lists in Python
In R we can do many operations on vectors for e.g. element by element addition, subtraction,
exponentation,product etc. as show
```{r cache=TRUE}
#R - Operations on vectors
a<- c(5,2,3,1,7)
b<- c(1,5,4,6,8)

#Element wise Addition
print(a+b)
#Element wise subtraction
print(a-b)
#Element wise product
print(a*b)
# Exponentiating the elements of a vector
print(a^2)
```

In Python to do this on lists we need to use the 'map' and the 'lambda' function as follows
```{python }
# Python - Operations on list
a =[5,2,3,1,7]
b =[1,5,4,6,8]

#Element wise addition with map & lambda
print(list(map(lambda x,y: x+y,a,b)))
#Element wise subtraction
print(list(map(lambda x,y: x-y,a,b)))
#Element wise product
print(list(map(lambda x,y: x*y,a,b)))
# Exponentiating the elements of a list
print(list(map(lambda x: x**2,a)))


```

However if we create ndarrays from lists then we can do the element wise addition,subtraction,product, etc. like R. Numpy is really a powerful module with many, many functions for matrix manipulations
```{python cache=TRUE}
import numpy as np
a =[5,2,3,1,7]
b =[1,5,4,6,8]
a=np.array(a)
b=np.array(b)
#Element wise addition
print(a+b)
#Element wise subtraction
print(a-b)
#Element wise product
print(a*b)
# Exponentiating the elements of a list
print(a**2)

```

###7. Getting the index of element
To determine the index of an element which satisifies a specific logical condition in R use 'which'. In the code below the index of  element which is equal to 1 is 4
```{r cache=TRUE}
# R - Which
a<- c(5,2,3,1,7)
print(which(a == 1))
```

In Python array we can use np.where to get the same effect. The index will be 3 as the index starts from 0
```{python cache=TRUE}
# Python - np.where
import numpy as np
a =[5,2,3,1,7]
a=np.array(a)
print(np.where(a==1))
```

###8. Data frames
R, by default comes with a set of in-built datasets. There are some datasets which come with the SkiKit- Learn package

```{r cache=TRUE}
# R 
# To check built datasets use
#data() - In R console or in R Studio
#iris - Don't print to console

```

We can use the in-built data sets that come with Scikit package
```{python cache=TRUE}
#Python
import sklearn as sklearn
import pandas as pd
from sklearn import datasets
# This creates a Sklearn bunch
data = datasets.load_iris()
# Convert to Pandas dataframe
iris = pd.DataFrame(data.data, columns=data.feature_names)
```

###9. Working with dataframes
With R you can work with dataframes directly. For more complex dataframe operations in R there are convenient packages like dplyr, reshape2 etc. For Python we need to use the Pandas package. Pandas is quite comprehensive in the list of things we can do with data frames
The most common operations on a dataframe are

- Check the size of the dataframe
- Take a look at the top 5 or bottom 5 rows of dataframe
- Check the content of the dataframe

####a.Size
In R use dim()
```{r cache=TRUE}
#R - Size
dim(iris)
```
For Python use .shape
```{python cache=TRUE}
#Python - size
import sklearn as sklearn
import pandas as pd
from sklearn import datasets
data = datasets.load_iris()
# Convert to Pandas dataframe
iris = pd.DataFrame(data.data, columns=data.feature_names)
iris.shape
```

####b. Top & bottom 5 rows of dataframe
To know the top and bottom rows of a data frame we use head() & tail as shown below for R
and Python
```{r cache=TRUE}
#R 
head(iris,5)
tail(iris,5)
```

```{python cache=TRUE}
#Python
import sklearn as sklearn
import pandas as pd
from sklearn import datasets
data = datasets.load_iris()
# Convert to Pandas dataframe
iris = pd.DataFrame(data.data, columns=data.feature_names)
print(iris.head(5))
print(iris.tail(5))
```

####c. Check the content of the dataframe
```{r cache=TRUE}
#R
summary(iris)
str(iris)
```


```{python}
#Python
import sklearn as sklearn
import pandas as pd
from sklearn import datasets
data = datasets.load_iris()
# Convert to Pandas dataframe
iris = pd.DataFrame(data.data, columns=data.feature_names)
print(iris.info())
```

####d. Check column names
```{r cache=TRUE}
#R
names(iris)
colnames(iris)
```


```{python cache=TRUE}
#Python
import sklearn as sklearn
import pandas as pd
from sklearn import datasets
data = datasets.load_iris()
# Convert to Pandas dataframe
iris = pd.DataFrame(data.data, columns=data.feature_names)
#Get column names
print(iris.columns)
```


####e. Rename columns
In R we can assign a vector to column names
```{r cache=TRUE}
#R
colnames(iris) <- c("lengthOfSepal","widthOfSepal","lengthOfPetal","widthOfPetal","Species")
colnames(iris)
```

In Python we can assign a list to <df>s.columns
```{python cache=TRUE}
#Python
import sklearn as sklearn
import pandas as pd
from sklearn import datasets
data = datasets.load_iris()
# Convert to Pandas dataframe
iris = pd.DataFrame(data.data, columns=data.feature_names)
iris.columns = ["lengthOfSepal","widthOfSepal","lengthOfPetal","widthOfPetal"]
print(iris.columns)
```


###f.Details of dataframe
```{python cache=TRUE}
#Python
import sklearn as sklearn
import pandas as pd
from sklearn import datasets
data = datasets.load_iris()
# Convert to Pandas dataframe
iris = pd.DataFrame(data.data, columns=data.feature_names)
print(iris.info())
```

####g. Subsetting dataframes
```{r cache=TRUE}
# R
#To subset a dataframe 'df' in R we use df[row,column] or df[row vector,column vector]
#df[row,column]
iris[3,4]
#df[row vector, column vector]
iris[2:5,1:3]
#If we omit the row vector, then it implies all rows or if we omit the column vector
# then implies all columns for that row
iris[2:5,]
# In R we can all specific columns by column names
iris$Sepal.Length[2:5]
```


```{python cache=TRUE}
#Python
# To select an entire row we use .iloc. The index can be used with the ':'. If 
# .iloc[start row: end row]. If start row is omitted then it implies the beginning of
# data frame, if end row is omitted then it implies all rows till end
#Python
import sklearn as sklearn
import pandas as pd
from sklearn import datasets
data = datasets.load_iris()
# Convert to Pandas dataframe
iris = pd.DataFrame(data.data, columns=data.feature_names)
print(iris.iloc[3])
print(iris[:5])
# In python we can select columns by column name as follows
print(iris['sepal length (cm)'][2:6])
#If you want to select more than 2 columns then you must use the double '[[]]' since the 
# index is a list itself
print(iris[['sepal length (cm)','sepal width (cm)']][4:7])
```


####h. Computing Mean, Standard deviation
```{r cache=TRUE}
#R 
#Mean
mean(iris$lengthOfSepal)
#Standard deviation
sd(iris$widthOfSepal)

```

```{python cache=TRUE}
#Python
#Mean
import sklearn as sklearn
import pandas as pd
from sklearn import datasets
data = datasets.load_iris()
# Convert to Pandas dataframe
iris = pd.DataFrame(data.data, columns=data.feature_names)
# Convert to Pandas dataframe
print(iris['sepal length (cm)'].mean())
#Standard deviation
print(iris['sepal width (cm)'].std())
```

####i. Boxplot
Boxplot can be produced in R using baseplot
```{r cache=TRUE}
#R
boxplot(iris$lengthOfSepal)
```

Matplotlib is a popular package in Python for plots
```{python cache=TRUE}
#Python
import sklearn as sklearn
import pandas as pd
import matplotlib.pyplot as plt
from sklearn import datasets
data = datasets.load_iris()
# Convert to Pandas dataframe
iris = pd.DataFrame(data.data, columns=data.feature_names)
import matplotlib.pyplot as plt
img=plt.boxplot(iris['sepal length (cm)'])
#plt.show(img)
```

####j. Scatter plot
```{r cache=TRUE}
#R
plot(iris$widthOfSepal,iris$lengthOfSepal)
```

```{python cache=TRUE}
#Python
import matplotlib.pyplot as plt
import sklearn as sklearn
import pandas as pd
from sklearn import datasets
data = datasets.load_iris()
# Convert to Pandas dataframe
iris = pd.DataFrame(data.data, columns=data.feature_names)
img=plt.scatter(iris['sepal width (cm)'],iris['sepal length (cm)'])
#plt.show(img)
```

####k. Read from csv file
```{r cache=TRUE}
#R
tendulkar= read.csv("tendulkar.csv",stringsAsFactors = FALSE,na.strings=c(NA,"-"))
#Dimensions of dataframe
dim(tendulkar)
names(tendulkar)
```

Use pandas.read_csv() for Python
```{python cache=TRUE}
#Python
import pandas as pd
#Read csv
tendulkar= pd.read_csv("tendulkar.csv",na_values=["-"])
print(tendulkar.shape)
print(tendulkar.columns)
```

####l. Clean the dataframe in R and Python. 
The following steps are done for R and Python

1.Remove rows with 'DNB'
2.Remove rows with 'TDNB'
3.Remove rows with absent
4.Remove the "* indicating not out
5.Remove incomplete rows with NA for R or NaN in Python
6.Do a scatter plot


```{r cache=TRUE}
#R
# Remove rows with 'DNB'
a <- tendulkar$Runs != "DNB"
tendulkar <- tendulkar[a,]
dim(tendulkar)
# Remove rows with 'TDNB'
b <- tendulkar$Runs != "TDNB"
tendulkar <- tendulkar[b,]

# Remove rows with absent
c <- tendulkar$Runs != "absent"
tendulkar <- tendulkar[c,]
dim(tendulkar)

# Remove the "* indicating not out
tendulkar$Runs <- as.numeric(gsub("\\*","",tendulkar$Runs))
dim(tendulkar)
# Select only complete rows - complete.cases()
c <- complete.cases(tendulkar)
#Subset the rows which are complete
tendulkar <- tendulkar[c,]
dim(tendulkar)
# Do some base plotting - Scatter plot
plot(tendulkar$BF,tendulkar$Runs)

```


```{python cache=TRUE}
#Python 
import pandas as pd
import matplotlib.pyplot as plt
#Read csv
tendulkar= pd.read_csv("tendulkar.csv",na_values=["-"])
print(tendulkar.shape)
# Remove rows with 'DNB'
a=tendulkar.Runs !="DNB"
tendulkar=tendulkar[a]
print(tendulkar.shape)
# Remove rows with 'TDNB'
b=tendulkar.Runs !="TDNB"
tendulkar=tendulkar[b]
print(tendulkar.shape)
# Remove rows with absent
c= tendulkar.Runs != "absent"
tendulkar=tendulkar[c]
print(tendulkar.shape)
# Remove the "* indicating not out
tendulkar.Runs= tendulkar.Runs.str.replace(r"[*]","")
#Select only complete rows - dropna()
tendulkar=tendulkar.dropna()
print(tendulkar.shape)
tendulkar.Runs = tendulkar.Runs.astype(int)
tendulkar.BF = tendulkar.BF.astype(int)
#Scatter plot
plt.scatter(tendulkar.BF,tendulkar.Runs)
```

####m. Chaining operations on dataframes
To chain a set of operations we need to use an R package like dplyr. Pandas does this 
The following operations are done on tendulkar data frame by dplyr for R and Pandas for Python below

1. Group by ground
2. Compute average runs in each ground
3. Arrange in descending order

```{r cache=TRUE}
#R
library(dplyr)
tendulkar1 <- tendulkar %>% group_by(Ground) %>% summarise(meanRuns= mean(Runs)) %>%
         arrange(desc(meanRuns))
head(tendulkar1,10)
```

```{python }
#Python
import pandas as pd
#Read csv
tendulkar= pd.read_csv("tendulkar.csv",na_values=["-"])
print(tendulkar.shape)
# Remove rows with 'DNB'
a=tendulkar.Runs !="DNB"
tendulkar=tendulkar[a]
# Remove rows with 'TDNB'
b=tendulkar.Runs !="TDNB"
tendulkar=tendulkar[b]
# Remove rows with absent
c= tendulkar.Runs != "absent"
tendulkar=tendulkar[c]
# Remove the "* indicating not out
tendulkar.Runs= tendulkar.Runs.str.replace(r"[*]","")

#Select only complete rows - dropna()
tendulkar=tendulkar.dropna()
tendulkar.Runs = tendulkar.Runs.astype(int)
tendulkar.BF = tendulkar.BF.astype(int)
tendulkar1= tendulkar.groupby('Ground').mean()['Runs'].sort_values(ascending=False)
print(tendulkar1.head(10))
```

###9. Functions
```{r}
product <- function(a,b){
  c<- a*b
  c
}
product(5,7)
```

```{python}
def product(a,b):
  c = a*b
  return c
  
print(product(5,7))


```

## Conclusion
Personally, I took to R, much like a 'duck takes to water'. I found the R syntax very simple and mostly intuitive. R packages like dplyr, ggplot2, reshape2, make the language quite irrestible. R is weakly typed and has only numeric and character types as opposed to the full fledged data types in Python. 

Python, has too many bells and whistles, which can be a little bewildering to the novice. It is possible that they may be useful as one becomes more experienced with the language. Also I found that installing Python packages sometimes gives errors with Python versions 2.7 or 3.6. This will leave you scrambling to google to find how to fix these problems. These can be quite frustrating. R on the other hand makes installing R packages a breeze.

Anyway, this is my current opinion, and like all opinions, may change in the course of time. Let's see!

I may write a follow up post with more advanced features of R and Python. So do keep checking!
Long live R! Viva la Python!